---
title: "Evaluate viral MGS protocols: Amicon centrifugal filters vs InnovaPrep Concentrating Pipette"
description: | 
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - R
  - qPCR
bibliography: ../../_references.bib
date: "`r format(Sys.time(), '%Y-%m-%d')`"
draft: false
output:
  distill::distill_article:
    self_contained: false
    dev: svg
    toc: true
    code_folding: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dpi = 300,
  include = TRUE
)
```

This experiment compares Amicon filters (30 kDa and 100 kDa) vs the InnovaPrep Concentrating Pipette (two tip types).

[Drive folder](https://drive.google.com/drive/folders/1ADu2YNofmC1iEt7CcvYEys-45bIg_70g)

```{r}
library(tidyverse)
library(googlesheets4)
library(fs)
library(here)

library(knitr)

library(broom)

# plotting helpers
library(cowplot)
library(patchwork)
library(ggbeeswarm)

theme_set(theme_cowplot())

# Okabe Ito color scheme with amber for yellow; see https://easystats.github.io/see/reference/scale_color_okabeito.html
colors_oi <- grDevices::palette.colors()  
colors_oi['yellow'] <- "#F5C710"

# Custon qPCR helpers
source('_functions.R')
```

# Data import

Metadata is stored in a Google Sheet,

```{r}
url <- 'https://docs.google.com/spreadsheets/d/1qBD6ZhDDtGBx1iYgcIJf6GZJhOZoS6B15vtvhJYdkoo'
meta_samples <- read_sheet(url, sheet = 'Samples', na = c("", "NA")) %>%
  glimpse
meta_targets <- read_sheet(url, sheet = 'Targets', na = c("", "NA")) %>%
  glimpse
```

Results are in CSV files,

```{r}
fns <- '_data' %>%
  dir_ls(recurse = TRUE, glob = '*_Results_*.csv')
fns %>% path_file
```

TODO: pull additional metadata from the comments in the CSV file

```{r}
results_raw <- tibble(file = fns) %>%
  mutate(.keep = 'unused',
    data = map(file, read_qpcr_results_csv)
  ) %>%
  unnest(data)
```

```{r}
results <- results_raw %>%
  rename(target_qpcr = target) %>%
  left_join(meta_samples, by = c('sample' = 'sample_qpcr')) %>%
  left_join(meta_targets, by = 'target_qpcr') %>%
  mutate(
    across(c(treatment_group, amicon_mwco, volume, method_short, filter), as.factor),
    cq = ifelse(is.na(cq), 40, cq),
    across(method_short, ~fct_relevel(.x, 'Amicon')),
    across(filter, ~fct_relevel(.x, '30 kDa')),
    method_filter = interaction(method_short, filter, drop = TRUE, sep = ': ', lex.order = TRUE)
  )
```

Note, in cases where the Cq value could not be determined, I set the value to 40.

```{r}
results %>% count(treatment_group, cq_status)
```

In this experiment, the Cq value was determined for all the non-control samples.

```{r}
results %>%
  count(target_qpcr, target)
```

Load amplification data,

```{r}
fns_amp <- '_data' %>%
  dir_ls(recurse = TRUE, glob = '*_Amplification Data_*.csv') %>%
  str_subset(negate = TRUE, 'Raw')

amp <- tibble(file = fns_amp) %>%
  mutate(.keep = 'unused',
    data = map(file, read_qpcr_amplification_csv)
  ) %>%
  unnest(data) %>%
  rename(target_qpcr = target) %>%
  left_join(results)
```

TODO: find a better way to import the amplification data and results at the same time

Get the baseline coordinates for plotting,

```{r}
baselines <- results %>%
  pivot_longer(
    cols = c(baseline_start, baseline_end),
    names_to = 'baseline_boundary',
    values_to = 'cycle_number',
    names_prefix = 'baseline_',
  ) %>%
  left_join(
    amp %>% select(well_position, cycle_number, rn, d_rn), 
    by = c('well_position', 'cycle_number')
  )
```


# Analysis

The source samples are either from the N or S sewer system.

The treatment group is determined by

#. Amicon vs InnovaPrep Concentrating Pipette (CP)
#. Filter pore size: Amicon MWCO (30 kDa or 100 kDa) or CP tip type (0.05um or 'Ultra')
#. Volume used (40 mL or 80 mL); only applies to CP

```{r}
results %>%
  count(treatment_group, method_short, filter, volume)
```

```{r, fig.dim = c(8, 5) * 1.3}
results %>%
  filter(!is.na(treatment_group)) %>%
  ggplot(aes(x = cq, y = method_filter, color = volume)) +
  theme(
    legend.position = 'bottom',
    panel.spacing = unit(2, 'char'),
    # panel.spacing.x = unit(4, 'char'),
  ) +
  scale_x_continuous(n.breaks = 4) +
  labs(
    y = 'Method : Filter',
    x = 'Cq',
    color = 'Volume (mL)',
  ) +
  facet_grid(sewer_system ~ target, scales = 'free') +
  scale_color_manual(values = colors_oi %>% unname) +
  geom_quasirandom()
```

```{r, fig.dim = c(7, 9) * 1.3}
results %>%
  filter(!is.na(treatment_group)) %>%
  ggplot(aes(x = cq, y = method_filter, color = volume)) +
  # theme_minimal_vgrid() +
  theme(
    legend.position = 'bottom',
    panel.spacing = unit(2, 'char'),
    # panel.spacing.x = unit(4, 'char'),
  ) +
  labs(
    y = 'Method : Filter',
    x = 'Cq',
    color = 'Volume (mL)',
  ) +
  # facet_grid(sewer_system ~ target, scales = 'free') +
  facet_wrap(
    target ~ sewer_system , scales = 'free_x', ncol = 2, 
    labeller = function(x) label_value(x, multi_line = FALSE)
  ) +
  scale_color_manual(values = colors_oi %>% unname) +
  geom_quasirandom()
```

Note that for the CP 0.05 um filter, it appears that the 80mL sample led to a lower concentration for the four naturally-occurring targets (Bacterial 16S, CrAssphage, Norovirus, and SARS2).
This is surprising and the only one of four 80mL samples where the measured concentrations were lower.
Might this indicate a mistake in the sample preparation?

The N Amicon 100 sample led to much lower measured concentrations than the N Amicon 30 sample for everything except SARS2.
Possibly there was a handling error or the Amicon filter was defective.

## Inspect SARS2 amplification curves

```{r, fig.dim = c(6,4)*1.5}
delta_rn_min <- 1e-3
ct_threshold <- results %>% filter(target == 'SARS-CoV-2') %>% pull(threshold) %>% unique
stopifnot(length(ct_threshold) == 1)

amp %>%
  filter(
    target == 'SARS-CoV-2',
    !is.na(treatment_group)
  ) %>%
  ggplot(aes(cycle_number, pmax(d_rn, delta_rn_min), color = method_filter)) +
  facet_grid(sewer_system ~ volume, scales = 'free') +
  scale_color_manual(values = colors_oi %>% unname) +
  scale_y_log10() +
  geom_line(aes(group = well)) +
  geom_hline(yintercept = ct_threshold, alpha = 0.3) +
  # scale_color_brewer(type = 'qual') +
  # geom_point(data = baselines, aes(shape = baseline_boundary), size = 3) +
  # scale_shape_manual(values = c(1, 4)) +
  labs(y = 'Delta Rn', x = 'Cycle', color = 'Target')
```


## Linear modeling

Fit linear model to assess the effect of conditions on the Cq value on each target.
Quick and unreliable analysis.

Measure effect on average Cq value of the concentration treatment relative to the Amicon 30 kDa treatment.
A positive effect corresponds to a decrease in concentration.

```{r}
fits <- results %>%
  filter(!is.na(treatment_group), volume == 40) %>%
  summarize(
    .by = c(target, method_short, method_filter, sewer_system),
    across(cq, mean)
  ) %>%
  nest(.by = target) %>%
  mutate(.keep = 'unused',
    fit = map(data, ~lm(data = .x, cq ~ sewer_system + method_filter)),
    # fit_tidy = map(fit, broom::tidy)
  )
fits_summary <- fits %>%
  mutate(.keep = 'unused',
    tidy = map(fit, broom::tidy)
  ) %>%
  unnest(tidy)
```

```{r}
fits_summary %>%
  filter(str_detect(term, 'method_filter')) %>%
  mutate(
    method_filter = str_replace(term, 'method_filter', '')
  ) %>%
  select(target, method_filter, estimate, std.error, p.value) 
```


```{r, fig.dim = c(8, 4)}
fits_summary %>%
  filter(
         str_detect(term, 'method_filter'),
         # str_detect(term, 'CP:'),
  ) %>%
  mutate(
    method_filter = str_replace(term, 'method_filter', '')
  ) %>%
  ggplot(aes(y = target, x = estimate, 
             xmin = estimate - 2 * std.error, 
             xmax = estimate + 2 * std.error, 
             color = method_filter
             )) +
  scale_color_manual(values = colors_oi %>% unname) +
  geom_vline(xintercept = 0, color = 'grey') +
  geom_pointrange(position = ggstance::position_dodgev(height = 0.3))
```


# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>
